---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data
**Not Shown: **Before I began, I unziped the activity.zip file into ./data/activity.csv  
Next, read the data in  
And, convert the date strings to date types  

### Code for reading in the dataset and/or processing the data
```{r}
thedata <- read.csv("./data/activity.csv", na.strings="NA")
thedata$date <- as.Date(as.character(thedata$date))
str(thedata)
```



## What is mean total number of steps taken per day?
Sum the steps by day  
Then, take a quick look  
And, the mean and median  

```{r}
totalsteps <- aggregate(steps ~ date, data=thedata, sum)
totalsteps 
```

### A histogram of the total number of steps taken each day
```{r}
hist(totalsteps$steps)
```

### Both the mean and median number of steps taken each day
```{r}
summary(totalsteps$steps)
```



## What is the average daily activity pattern?
Take the mean of steps by interval  
And see which interval has the highest average  
Then, plot them  

```{r}
averagesteps <- aggregate(steps ~ interval, data=thedata, FUN="mean")
averagesteps[averagesteps$steps==max(averagesteps$steps),] # get the entry at max
```

### A time series plot of the average number of steps taken (averaged across all days) versus the 5-minute intervals
```{r}
plot(averagesteps, type="l")
```

###  The 5-minute interval that, on average, contains the maximum number of steps
Interval number `r averagesteps[averagesteps$steps==max(averagesteps$steps),]$interval` has the highest average steps, at `r averagesteps[averagesteps$steps==max(averagesteps$steps),]$steps`!


## Imputing missing values
There are `r sum(is.na(thedata$steps))` missing values in the steps column.  
There are `r sum(is.na(thedata$date))` missing values in the date column.  
There are `r sum(is.na(thedata$interval))` missing values in the interval column.  

### Describe and show with code a strategy for imputing missing data
Next we'll take a copy of the data   
And, replace the NA steps counts with the average for that interval for all days  

```{r}
thedata.nona <- thedata

thedata.nona$steps <- apply(thedata.nona, 1, function(x) { 
  if(is.na(x[1])) { subset(averagesteps,interval==as.integer(x[3]))[1,2] } 
  else { as.integer(x[1]) }
})

str(thedata.nona)
```

Let's look at the histograms side by side  

### A histogram of the total number of steps taken each day after missing values were imputed
```{r}
totalsteps.nona <- aggregate(steps ~ date, data=thedata.nona, sum)

par(mfcol=c(1,2))
hist(totalsteps$steps)
hist(totalsteps.nona$steps)
```

And the summaries  

```{r}
summary(totalsteps$steps)
summary(totalsteps.nona$steps)
```

Imputing the missing value by using the interval mean seems to have had a minimal impact on the median and mean values for the average daily totals.



## Are there differences in activity patterns between weekdays and weekends?
Let's add a weekday factor  
Then a weekend factor  

```{r}
thedata.nona$weekday <- as.factor(weekdays(thedata.nona$date))
thedata.nona$weekpart <- as.factor(thedata.nona$weekday %in% c("Saturday", "Sunday"))
levels(thedata.nona$weekpart) <- c("Weekday","Weekend")
```

Now take the mean for weekdays and weekend  
And graph the two  

### Plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
averagesteps.weekpart <- aggregate(steps ~ interval+weekpart, data=thedata.nona,FUN="mean")
library(ggplot2)
qplot(interval,steps,data=averagesteps.weekpart, geom="line", color=weekpart)
```

~~~~
Josh Blacksher
~~~~
